Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 4.dev

### 2025-11-16 main.tex
Change all citep and citet to cite.

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage (\textgreater80\%) of reported serious injuries in The Netherlands
and values range between 50\% and 95\% in other surveyed
countries~\citep{Wegman2024}. \citet{Utriainen2023} report that slippery
surfaces, simple loss of control, and engagement with a curb, pothole, or
bollard are most the common fall types in the seven reported countries. Most of
these types involve perturbations that result in loss of balance. At low
speeds, from start up to typical cruising speeds, bicycles are not self-stable
and can be challenging for the rider to balance. Low speed crashes may be
reduced if the bicycle was self-stable at these speeds by relieving the rider
of some of their necessary control activity. Bicycles can be mechanically
modified to lower the speeds at which they are self-stable~\citep{Astrom2005}
and since the 1980s it has been known that applying a motor actuated steering
torque proportional to the vehicle's roll angular rate~\footnote{``Roll'' is
defined herein as per the SAE J670 vehicle standards.} can stabilize a
single-track vehicle down to very low speeds~\citep{Ruijs1986}. If automatic
control of steering can stabilize a bicycle, it may reduce the control required
from the rider to successfully manage balancing tasks much like natural high
speed self-stability already does. We have developed a balance-assisting
bicycle, Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack first introduced in \citet{Alizadehsaravi2023}.}
  \label{fig:balance-assist-bicycle}
\end{figure}

Riders fall when their control authority is not able to maintain a stable
vehicle state. There are many real world scenarios that can put the vehicle
into an uncontrollable state. External forces applied to the vehicle or rider
are one such scenario type and natural examples include wind gusts, handlebars
colliding with a neighbor's, a bag swinging from the handlebar, or simply
hitting a bump in the road. To assess our balance assisting bicycle, we subject
the rider to perturbations at the handlebar, which can easily cause a rider to
fall. From a mechanics perspective, it is not particularly important how or
where the perturbation is applied but that it results in forcing the
bicycle-rider system into an unrecoverable state, as all forces acting on a
dynamical system can be resolved into an equivalent force and torque of a
couple at any location in the system~\citep{Kane1985}. We chose to perturb the
handlebar because of this superposition principle and that it was both
practical and required a low force to cause a fall.

@@

\subsection{Technical Background}
%
During the early years of developments in automatic control,
\citet{Whipple1899} not only derived the correct equations of motion of the
bicycle \citep{Meijaard2007} but realized and showed that roll motion feedback
can stabilize a bicycle. Much later, attempts at automatic roll stabilization
of single-track vehicles began in earnest after predictive motorcycle models
were developed and refined throughout the 1970s. \Citet{vanZytveld1975} was
influenced by Whipple's work and seems to be the first to attempt to
robotically stabilize a small motorbike with a controlled inverted pendulum
that mimicked rider lean, but he was not successful in demonstrating what his
theoretic control model correctly predicted. In his model, he recognized that
feedback of the vehicle roll angle and angular rate was essential to stabilize
the vehicle. It was not until the early 1980s that \citet{Nagai1983}
successfully demonstrated balancing a robotic bicycle on a treadmill with both
steering control and a laterally moving mass. \citet{Ruijs1986} followed this
breakthrough by demonstrating an automatically balanced motorcycle and they did
so solely with a steering motor. Ruijs and Pacejka showed that steer torque
driven by roll angle feedback stabilizes the capsize mode, by roll angular rate
feedback stabilizes the weave mode, and by steer angular rate feedback
stabilizes the wobble mode.~\footnote{These motorcycle (and bicycle) eigenmodes
are defined in \citep{Sharp1971}.} They also showed how the control gains must
change with respect to vehicle speed for favorable control across all speeds.
Thus roll motion feedback enables the simplest controller that can stabilize a
single-track vehicle above a minimum speed when one is not concerned with
wobble instabilities. But Ruijs and Pacejka's work was not particularly
concerned with low speed stability and their vehicle was fully automatic, i.e
no human rider was involved.

Many more automatically balanced single-track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in human balancing and what effect it may have towards
single-actor falls. Most of these robotic bicycle and motorcycle designers did
not intend for a human rider to also control the stabilized vehicle.
Nevertheless, an automatically stabilized bicycle can be controlled by a human
rider if the motor controlled steer torque and the rider applied steer torque
act in parallel. The effect of this automatic control gives the ability to
effectively change the human-controlled plant dynamics, up to some limits. In
our prior study~\citep{Alizadehsaravi2023}, we demonstrate reduced motion
during distractions and light perturbations due to the balance assist system
but \citet{Hanakam2023} recently showed rider dissatisfaction with the
stabilization of a similar vehicle so the overall possible benefits are not yet
definitively established.

The design of our balance assist system relies on the linear Carvallo-Whipple
bicycle model~\citep{Carvallo1899,Whipple1899} which is the simplest bicycle
model that exhibits both the non-minimum phase behavior ``countersteering'' and
self-stability. The bicycle model is well suited for showing the effect of a
roll motion feedback driven steer motor on the dynamics. This model is
equivalently valid for on-road or treadmill riding~\citep{Kooijman2009a}, which
have the same fundamental dynamics. The linear version of this model can be
described by the fourth order state space equations:
%
\begin{align}
  \dot{\vec{x}} = \mathbf{A} \vec{x} +
  \mathbf{B}
  \vec{u}
  \textrm{ where }
  \vec{x} = \begin{bmatrix} \phi \\ \dot{\phi} \\ \delta \\ \dot{\delta} \end{bmatrix}
  \textrm{ and }
  \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
  \textrm{.}
\end{align}
%
The states \(\vec{x}\) are the roll angle \(\phi\) and steer angle \(\delta\)
along with their time derivatives and the inputs \(\vec{u}\) are roll torque
\(T_\phi\) and steer torque \(T_\delta\). The state matrix \(\mathbf{A}\) is a
function of the equilibrium forward speed \(v\). It and the input matrix
\(\mathbf{B}\) are otherwise populated with expressions that are functions of
the geometric and inertial parameters of the nonholonomic multibody system made
up of four rigid bodies: two wheels, front frame, and rear frame.

@@

The state matrix \(\mathbf{A}\) being a function of the equilibrium speed \(v\)
means the control gain \(k_{\dot{\phi}}\) can be selected such that the
eigenvalues of \(\left(\mathbf{A} - \vec{B}_\delta\left[0 \quad k_{\dot{\phi}}
\quad 0 \quad 0\right] \right)\) have negative real parts for \(v_{min} < v <
v_\textrm{capsize}\) where \(v_{min}\) is the lowest stable speed given
\(k_{\dot{\phi}}\) and \(v_\textrm{capsize}\) is the speed at which the
uncontrolled bicycle's capsize mode goes unstable. By gain scheduling with
respect to \(v\), the speed range where the bicycle is stable can be maximized
within any physical actuator magnitude and bandwidth limits.
\citet{Schwab2008} elaborate on some of the possibilities in scheduling the
gains for such a controller for a bicycle and show that linear scheduling with
respect to speed can give satisfactory stability for a low speed range. We use
this simple feedback principle as the basis for our balance assist controller.

@@

\subsection{Balance Assist Control}
%
We use a forward speed \(v\) gain scheduled proportional roll angular rate
feedback controller to stabilize the bicycle. In the speed range tested, the
commanded steer torque \(T^\textrm{m}_\delta\) from the steer motor follows the
control law
%
\begin{align}
  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = \kappa(v_{stable} - v)\dot{\phi}
  \label{eq:implemented-controller}
\end{align}
%
where \(v_{\textrm{stable}} = 4.7~\si{\mps}\) is approximately the average
stable speed predicted from the open-loop bicycle rigid-rider dynamics.  We use
the gain factor \(\kappa=3.9\) (low) and \(\kappa=5.2\) (high) to tune the gain
magnitude during the experiments. Figure~\ref{fig:eigenvalues} shows the
stabilization effect of the controller via the eigenmodes of the linear system.
Geometric and inertial parameters for these plots were estimated with the
methods presented in \citet{Moore2012} and software packages
BicycleParameters~1.1.1~\citep{Moore2024} and Yeadon~1.5.0~\citep{Dembia2015}
and are shown in Appendix~\ref{sec:physical-parameters}.

For the 6~\si{\kph} experiment, scaling the proportional feedback gain linearly
with respect to speed stabilizes the normally unstable weave mode of the
bicycle down to 3.4~\si{\kph} for the riderless bicycle and 8.0~\si{\kph} for
the ridden bicycle. For the 10~\si{\kph} experiment, it stabilizes the normally
unstable weave mode of the bicycle down to 3.8~\si{\kph} for the riderless
bicycle and 9.7~\si{\kph} for the ridden bicycle as shown in
Figure~\ref{fig:eigenvalues}. Notably, the bicycle is not stable at 6~\si{\kph}
but the time constant of the unstable eigenvalue is increased by a factor of 8
and at 10~\si{\kph} the system is just stable. Figure~\ref{fig:eigenvalues}
also shows the identified weave mode captured using the methods of
Kooijman~et~al.~\citep{Kooijman2009a} and gives an indication of the
performance of the system, with the model predicting the weave mode well the
region of the tested speeds.
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row), controlled with \(\kappa=3.9\) (middle
    row), and controlled with \(\kappa=5.2\) (lower row) root locus of the
    eigenvalue components (real: solid, imaginary: dashed) plotted versus speed
    for the linear Carvallo-Whipple bicycle model without the inertial effect
    of a rigid rider (left) and with a rigid rider (right). Vertical dotted and
    dotted-dashed lines indicate the two speeds we perform experiments at:
    6~\si{\kph} and 10~\si{\kph}. The grey shaded region is the linear stable
    speed range. Black stars are experimentally identified weave mode decay and
    frequency.}
  \label{fig:eigenvalues}
\end{figure}

\subsection{Perturbations}
%
We apply longitudinal forces to the ends of each handlebar from earth-fixed
posts using an adapted Bump'Em system~\citep{Tan2020} which we arrange with
four EC-90 Flat motors (Maxon Group, Switzerland) working in cooperation,
Figure~\ref{fig:setup-diagram}. The four motors are programmed to apply a light
force at all times to keep the ropes taught and to track a commanded force
profile using a PID controller running on an Arduino Mega 2560 microprocessor
(Arduino, Italy). We measure the force applied by each motor at the handlebar
via four inline S-style load cells (Bosche GmbH, Damme, Germany). All were
rated for 250~\si{\newton} except the left rear load cell which was rated for
500~\si{\newton}. The commanded force profiles are designed to apply an
external pulsive torque to the front frame at magnitudes varying from
\SIrange{16}{160}{\newton\meter}. The four motors are arranged at the four
corners of a 1~\si{\meter} wide by 2~\si{\meter} long treadmill (Bonte
Technology B.V., Zwolle, The Netherlands) that can reach speeds up to
18~\si{\kph}. The general design of this perturbation system is described in
detail in Van De Velde's MSc thesis~\citep{vandeVelde2022a} and the physical
arrangement is shown in Figure~\ref{fig:participant-in-set-up}.  Our
modifications relative to Van De Velde's design include simplifying the Bump'Em
motor controller with an inexpensive microcontroller and the use of a simpler
non-active safety harness.
%
% NOTE : I get a compile error without the \protect calls here, not sure why.
\begin{figure}
  \centering
  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
    system.\protect\label{fig:setup-diagram}}
    {\includegraphics[height=65mm]{figures/setup-diagram.png}}
  \subcaptionbox{A participant on the bicycle in the safety harness with the
    Bump'Em motor lines attached to the ends of the
    handlebars.\protect\label{fig:participant-in-set-up}}
    {\includegraphics[height=65mm]{figures/participant-in-set-up.png}}
  \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
    perturbations and the various system components.}
  \label{fig:experimental-setup}
\end{figure}

@@

Participants wore a helmet and they were attached to the ceiling via a fall
safety harness, Figure~\ref{fig:participant-in-set-up}. The harness allowed
natural free movement pre-fall. The participants practiced falling in the
harness at least once and then practiced riding on the treadmill until they
indicated they were comfortable enough to have perturbations applied. For most,
this was less than a 10~\si{\minute} warm up. We then asked the rider to ride
for 90~\si{\second}, while attempting to maintain the location of their front
wheel on the center line of the treadmill as a baseline measure before the
perturbations. We then applied perturbations in random directions (clockwise or
counter-clockwise), starting at 20~\si{\newton} and increasing the magnitude by
30~\si{\newton} until the participants fell. We defined a ``fall'' on the
treadmill by two criteria: 1) the rider removes their foot from the pedal and
places it on the ground or 2) the bicycle wheel exceeds the width of the
treadmill belt. These two fall criteria are not strictly the same and the
latter is more likely to happen at higher speeds. We informed the participants
beforehand that we would stop the treadmill in these situations. These were
the two actions which necessitated stopping the treadmill belt and the simulate
real fall modes. Placing a foot down on the fall side of the bicycle is a
natural reaction when steering or other body movements are not sufficient to
keep the bicycle upright, as shown in \citep{Gildea2024}. Exiting the width of
the treadmill is akin to exiting the width of a curb-less bicycle path and
arises when the rider's control actions are not sufficient to correct the
heading of the vehicle in sufficient time.
Figure~\ref{fig:perturbation-sequence} shows an example resulting motion from a
perturbation. We logged the force magnitude that caused the first fall to
characterize that participant's \emph{perturbation resistance threshold}.
%
\begin{figure}
  \centering
  \subcaptionbox{Before perturbation}{\includegraphics[width=60mm]{figures/start.png}}
  \subcaptionbox{During perturbation}{\includegraphics[width=60mm]{figures/during.png}}
  \subcaptionbox{After perturbation}{\includegraphics[width=60mm]{figures/after.png}}
  \subcaptionbox{Recovery from perturbation}{\includegraphics[width=60mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

@@

Based on findings from measuring riders without balance assist in
\citep{Haitjema2023a}, we calculate several variables that we hypothesize may
influence fall probability. We use the angular impulse \(L\) of the
perturbation forces over a 0.3~\si{\second} duration to characterize the
magnitude of delivered perturbation. The duration is selected based on the
duration of the commanded perturbation force and is calculated as follows:
%
\begin{align}
  L =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left(F_r + F_l\right) dt
  =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left[\left(F_{rf} -
    F_{rr}\right) + \left(F_{lf} - F_{lr}\right)\right] dt
  \label{eq:angular-impulse}
\end{align}
%
where \(F_r\) and \(F_l\)  is the total force applied on the right and left
handlebar ends, respectively which are the sum of the rear and front load cell
readings \(F_{rr}\) and \(F_{rf}\), for example. The handlebar length is given
as \(l\) in Equation \ref{eq:angular-impulse}. We use angular impulse instead
of peak torque to normalize for the duration of the applied perturbation to
capture the total effect of the perturbation pulse.

@@

Before fitting the model, we scale each independent variable \(x_{ij}^k\) such
that they have a mean of zero and a standard deviation of one by cluster-mean
centering with clusters being an individual participant, as recommended by
\citet{Enders2007}. The clusters are chosen as all data from an individual
participant because we are only interested in the association between the state
of the balance assist system and the outcome of the perturbation for that
participant, not how participants perform relative to each other. We expected
there to be variation between participants in how well they are able to resist
the perturbation; however, this was not true. If a random effect is included in
the model for each participant, the intraclass correlation (ICC) is less that
3\% for both the 6~\si{\kph} and 10~\si{\kph} datasets, so we utilize a simple
single-level logistic regression model without the random effect instead of a
multilevel model. This tracks with the independent data collected in
\citet{Haitjema2023a} which also had negligible participant variation. This
left us with angular impulse, perturbation order, balance assist state, and
roll \& steer angles at the time of perturbation as independent variables. We
also include interaction effects between the balance assist state and the other
four variables. We divide the analysis into two separate model evaluations, one
for the 6~\si{\kph}, \(\kappa=5.2\) trials and one for the 10~\si{\kph},
\(\kappa=3.9\) trials and we evaluate our hypothesis for each set of data.
Statistics were computed with R version 4.1.3~\citep{RCoreTeam2021} the lme4
version 1.1~\citep{Bates2015} package.

@@

\subsection{Stability and Human Controlled Plant Dynamics}
%
The linear Carvallo-Whipple model indicates that the steer controller
stabilizes the bicycle-rider system, but this model assumes the rider's hands
are not connected to the handlebars and that they clamp their body as rigidly
as possible to the rear frame. In reality, the system's behavior is likely more
akin to a marginally stable or an easily controllable unstable system due to
the various un-modeled effects. Our system may not result in a definitely
stable system, i.e. cannot fall, but having plant eigenvalues with very small
unstable eigenvalue real parts correlates to ease of control~\citep{Hess2012}
which can be beneficial for avoiding falls.

>>>>

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage (\textgreater80\%) of reported serious injuries in The Netherlands
and values range between 50\% and 95\% in other surveyed
countries~\cite{Wegman2024}. Utriainen, O’Hern, and
Pöllänen~\cite{Utriainen2023} report that slippery surfaces, simple loss of
control, and engagement with a curb, pothole, or bollard are most the common
fall types in the seven reported countries. Most of these types involve
perturbations that result in loss of balance. At low speeds, from start up to
typical cruising speeds, bicycles are not self-stable and can be challenging
for the rider to balance. Low speed crashes may be reduced if the bicycle was
self-stable at these speeds by relieving the rider of some of their necessary
control activity. Bicycles can be mechanically modified to lower the speeds at
which they are self-stable~\cite{Astrom2005} and since the 1980s it has been
known that applying a motor actuated steering torque proportional to the
vehicle's roll angular rate~\footnote{``Roll'' is defined herein as per the SAE
J670 vehicle standards.} can stabilize a single-track vehicle down to very low
speeds~\cite{Ruijs1986}. If automatic control of steering can stabilize a
bicycle, it may reduce the control required from the rider to successfully
manage balancing tasks much like natural high speed self-stability already
does. We have developed a balance-assisting bicycle,
Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack first introduced in \cite{Alizadehsaravi2023}.}
  \label{fig:balance-assist-bicycle}
\end{figure}

Riders fall when their control authority is not able to maintain a stable
vehicle state. There are many real world scenarios that can put the vehicle
into an uncontrollable state. External forces applied to the vehicle or rider
are one such scenario type and natural examples include wind gusts, handlebars
colliding with a neighbor's, a bag swinging from the handlebar, or simply
hitting a bump in the road. To assess our balance assisting bicycle, we subject
the rider to perturbations at the handlebar, which can easily cause a rider to
fall. From a mechanics perspective, it is not particularly important how or
where the perturbation is applied but that it results in forcing the
bicycle-rider system into an unrecoverable state, as all forces acting on a
dynamical system can be resolved into an equivalent force and torque of a
couple at any location in the system~\cite{Kane1985}. We chose to perturb the
handlebar because of this superposition principle and that it was both
practical and required a low force to cause a fall.

@@

\subsection{Technical Background}
%
During the early years of developments in automatic control,
Whipple~\cite{Whipple1899} not only derived the correct equations of motion of
the bicycle \cite{Meijaard2007} but realized and showed that roll motion
feedback can stabilize a bicycle. Much later, attempts at automatic roll
stabilization of single-track vehicles began in earnest after predictive
motorcycle models were developed and refined throughout the 1970s. Van
Zytveld~\cite{vanZytveld1975} was influenced by Whipple's work and seems to be
the first to attempt to robotically stabilize a small motorbike with a
controlled inverted pendulum that mimicked rider lean, but he was not
successful in demonstrating what his theoretic control model correctly
predicted. In his model, he recognized that feedback of the vehicle roll angle
and angular rate was essential to stabilize the vehicle. It was not until the
early 1980s that Nagai~\cite{Nagai1983} successfully demonstrated balancing a
robotic bicycle on a treadmill with both steering control and a laterally
moving mass. Ruijs and Pacejka~\cite{Ruijs1986} followed this breakthrough by
demonstrating an automatically balanced motorcycle and they did so solely with
a steering motor. Ruijs and Pacejka showed that steer torque driven by roll
angle feedback stabilizes the capsize mode, by roll angular rate feedback
stabilizes the weave mode, and by steer angular rate feedback stabilizes the
wobble mode.~\footnote{These motorcycle (and bicycle) eigenmodes are defined in
\cite{Sharp1971}.} They also showed how the control gains must change with
respect to vehicle speed for favorable control across all speeds.  Thus roll
motion feedback enables the simplest controller that can stabilize a
single-track vehicle above a minimum speed when one is not concerned with
wobble instabilities. But Ruijs and Pacejka's work was not particularly
concerned with low speed stability and their vehicle was fully automatic, i.e
no human rider was involved.

Many more automatically balanced single-track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in human balancing and what effect it may have towards
single-actor falls. Most of these robotic bicycle and motorcycle designers did
not intend for a human rider to also control the stabilized vehicle.
Nevertheless, an automatically stabilized bicycle can be controlled by a human
rider if the motor controlled steer torque and the rider applied steer torque
act in parallel. The effect of this automatic control gives the ability to
effectively change the human-controlled plant dynamics, up to some limits. In
our prior study~\cite{Alizadehsaravi2023}, we demonstrate reduced motion during
distractions and light perturbations due to the balance assist system but
Hanakam, Wehner, and Wrede~\cite{Hanakam2023} recently showed rider
dissatisfaction with the stabilization of a similar vehicle so the overall
possible benefits are not yet definitively established.

The design of our balance assist system relies on the linear Carvallo-Whipple
bicycle model~\cite{Carvallo1899,Whipple1899} which is the simplest bicycle
model that exhibits both the non-minimum phase behavior ``countersteering'' and
self-stability. The bicycle model is well suited for showing the effect of a
roll motion feedback driven steer motor on the dynamics. This model is
equivalently valid for on-road or treadmill riding~\cite{Kooijman2009a}, which
have the same fundamental dynamics. The linear version of this model can be
described by the fourth order state space equations:
%
\begin{align}
  \dot{\vec{x}} = \mathbf{A} \vec{x} +
  \mathbf{B}
  \vec{u}
  \textrm{ where }
  \vec{x} = \begin{bmatrix} \phi \\ \dot{\phi} \\ \delta \\ \dot{\delta} \end{bmatrix}
  \textrm{ and }
  \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
  \textrm{.}
\end{align}
%
The states \(\vec{x}\) are the roll angle \(\phi\) and steer angle \(\delta\)
along with their time derivatives and the inputs \(\vec{u}\) are roll torque
\(T_\phi\) and steer torque \(T_\delta\). The state matrix \(\mathbf{A}\) is a
function of the equilibrium forward speed \(v\). It and the input matrix
\(\mathbf{B}\) are otherwise populated with expressions that are functions of
the geometric and inertial parameters of the nonholonomic multibody system made
up of four rigid bodies: two wheels, front frame, and rear frame.

@@

The state matrix \(\mathbf{A}\) being a function of the equilibrium speed \(v\)
means the control gain \(k_{\dot{\phi}}\) can be selected such that the
eigenvalues of \(\left(\mathbf{A} - \vec{B}_\delta\left[0 \quad k_{\dot{\phi}}
\quad 0 \quad 0\right] \right)\) have negative real parts for \(v_{min} < v <
v_\textrm{capsize}\) where \(v_{min}\) is the lowest stable speed given
\(k_{\dot{\phi}}\) and \(v_\textrm{capsize}\) is the speed at which the
uncontrolled bicycle's capsize mode goes unstable. By gain scheduling with
respect to \(v\), the speed range where the bicycle is stable can be maximized
within any physical actuator magnitude and bandwidth limits.  Schwab, Kooijman,
and Meijaard~\cite{Schwab2008} elaborate on some of the possibilities in
scheduling the gains for such a controller for a bicycle and show that linear
scheduling with respect to speed can give satisfactory stability for a low
speed range. We use this simple feedback principle as the basis for our balance
assist controller.

@@

\subsection{Balance Assist Control}
%
We use a forward speed \(v\) gain scheduled proportional roll angular rate
feedback controller to stabilize the bicycle. In the speed range tested, the
commanded steer torque \(T^\textrm{m}_\delta\) from the steer motor follows the
control law
%
\begin{align}
  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = \kappa(v_{stable} - v)\dot{\phi}
  \label{eq:implemented-controller}
\end{align}
%
where \(v_{\textrm{stable}} = 4.7~\si{\mps}\) is approximately the average
stable speed predicted from the open-loop bicycle rigid-rider dynamics.  We use
the gain factor \(\kappa=3.9\) (low) and \(\kappa=5.2\) (high) to tune the gain
magnitude during the experiments. Figure~\ref{fig:eigenvalues} shows the
stabilization effect of the controller via the eigenmodes of the linear system.
Geometric and inertial parameters for these plots were estimated with the
methods presented in \cite{Moore2012} and software packages
BicycleParameters~1.1.1~\cite{Moore2024} and Yeadon~1.5.0~\cite{Dembia2015} and
are shown in Appendix~\ref{sec:physical-parameters}.

For the 6~\si{\kph} experiment, scaling the proportional feedback gain linearly
with respect to speed stabilizes the normally unstable weave mode of the
bicycle down to 3.4~\si{\kph} for the riderless bicycle and 8.0~\si{\kph} for
the ridden bicycle. For the 10~\si{\kph} experiment, it stabilizes the normally
unstable weave mode of the bicycle down to 3.8~\si{\kph} for the riderless
bicycle and 9.7~\si{\kph} for the ridden bicycle as shown in
Figure~\ref{fig:eigenvalues}. Notably, the bicycle is not stable at 6~\si{\kph}
but the time constant of the unstable eigenvalue is increased by a factor of 8
and at 10~\si{\kph} the system is just stable. Figure~\ref{fig:eigenvalues}
also shows the identified weave mode captured using the methods of
Kooijman~et~al.~\cite{Kooijman2009a} and gives an indication of the performance
of the system, with the model predicting the weave mode well the region of the
tested speeds.
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row), controlled with \(\kappa=3.9\) (middle
    row), and controlled with \(\kappa=5.2\) (lower row) root locus of the
    eigenvalue components (real: solid, imaginary: dashed) plotted versus speed
    for the linear Carvallo-Whipple bicycle model without the inertial effect
    of a rigid rider (left) and with a rigid rider (right). Vertical dotted and
    dotted-dashed lines indicate the two speeds we perform experiments at:
    6~\si{\kph} and 10~\si{\kph}. The grey shaded region is the linear stable
    speed range. Black stars are experimentally identified weave mode decay and
    frequency.}
  \label{fig:eigenvalues}
\end{figure}

\subsection{Perturbations}
%
We apply longitudinal forces to the ends of each handlebar from earth-fixed
posts using an adapted Bump'Em system~\cite{Tan2020} which we arrange with four
EC-90 Flat motors (Maxon Group, Switzerland) working in cooperation,
Figure~\ref{fig:setup-diagram}. The four motors are programmed to apply a light
force at all times to keep the ropes taught and to track a commanded force
profile using a PID controller running on an Arduino Mega 2560 microprocessor
(Arduino, Italy). We measure the force applied by each motor at the handlebar
via four inline S-style load cells (Bosche GmbH, Damme, Germany). All were
rated for 250~\si{\newton} except the left rear load cell which was rated for
500~\si{\newton}. The commanded force profiles are designed to apply an
external pulsive torque to the front frame at magnitudes varying from
\SIrange{16}{160}{\newton\meter}. The four motors are arranged at the four
corners of a 1~\si{\meter} wide by 2~\si{\meter} long treadmill (Bonte
Technology B.V., Zwolle, The Netherlands) that can reach speeds up to
18~\si{\kph}. The general design of this perturbation system is described in
detail in Van De Velde's MSc thesis~\cite{vandeVelde2022a} and the physical
arrangement is shown in Figure~\ref{fig:participant-in-set-up}.  Our
modifications relative to Van De Velde's design include simplifying the Bump'Em
motor controller with an inexpensive microcontroller and the use of a simpler
non-active safety harness.
%
% NOTE : I get a compile error without the \protect calls here, not sure why.
\begin{figure}
  \centering
  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
    system.\protect\label{fig:setup-diagram}}
    {\includegraphics[height=65mm]{figures/setup-diagram.png}}
  \subcaptionbox{A participant on the bicycle in the safety harness with the
    Bump'Em motor lines attached to the ends of the
    handlebars.\protect\label{fig:participant-in-set-up}}
    {\includegraphics[height=65mm]{figures/participant-in-set-up.png}}
  \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
    perturbations and the various system components.}
  \label{fig:experimental-setup}
\end{figure}

@@

Participants wore a helmet and they were attached to the ceiling via a fall
safety harness, Figure~\ref{fig:participant-in-set-up}. The harness allowed
natural free movement pre-fall. The participants practiced falling in the
harness at least once and then practiced riding on the treadmill until they
indicated they were comfortable enough to have perturbations applied. For most,
this was less than a 10~\si{\minute} warm up. We then asked the rider to ride
for 90~\si{\second}, while attempting to maintain the location of their front
wheel on the center line of the treadmill as a baseline measure before the
perturbations. We then applied perturbations in random directions (clockwise or
counter-clockwise), starting at 20~\si{\newton} and increasing the magnitude by
30~\si{\newton} until the participants fell. We defined a ``fall'' on the
treadmill by two criteria: 1) the rider removes their foot from the pedal and
places it on the ground or 2) the bicycle wheel exceeds the width of the
treadmill belt. These two fall criteria are not strictly the same and the
latter is more likely to happen at higher speeds. We informed the participants
beforehand that we would stop the treadmill in these situations. These were the
two actions which necessitated stopping the treadmill belt and the simulate
real fall modes. Placing a foot down on the fall side of the bicycle is a
natural reaction when steering or other body movements are not sufficient to
keep the bicycle upright, as shown in \cite{Gildea2024}. Exiting the width of
the treadmill is akin to exiting the width of a curb-less bicycle path and
arises when the rider's control actions are not sufficient to correct the
heading of the vehicle in sufficient time.
Figure~\ref{fig:perturbation-sequence} shows an example resulting motion from a
perturbation. We logged the force magnitude that caused the first fall to
characterize that participant's \emph{perturbation resistance threshold}.
%
\begin{figure}
  \centering
  \subcaptionbox{Before perturbation}{\includegraphics[width=60mm]{figures/start.png}}
  \subcaptionbox{During perturbation}{\includegraphics[width=60mm]{figures/during.png}}
  \subcaptionbox{After perturbation}{\includegraphics[width=60mm]{figures/after.png}}
  \subcaptionbox{Recovery from perturbation}{\includegraphics[width=60mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

@@

Based on findings from measuring riders without balance assist in
\cite{Haitjema2023a}, we calculate several variables that we hypothesize may
influence fall probability. We use the angular impulse \(L\) of the
perturbation forces over a 0.3~\si{\second} duration to characterize the
magnitude of delivered perturbation. The duration is selected based on the
duration of the commanded perturbation force and is calculated as follows:
%
\begin{align}
  L =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left(F_r + F_l\right) dt
  =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left[\left(F_{rf} -
    F_{rr}\right) + \left(F_{lf} - F_{lr}\right)\right] dt
  \label{eq:angular-impulse}
\end{align}
%
where \(F_r\) and \(F_l\)  is the total force applied on the right and left
handlebar ends, respectively which are the sum of the rear and front load cell
readings \(F_{rr}\) and \(F_{rf}\), for example. The handlebar length is given
as \(l\) in Equation \ref{eq:angular-impulse}. We use angular impulse instead
of peak torque to normalize for the duration of the applied perturbation to
capture the total effect of the perturbation pulse.

@@

Before fitting the model, we scale each independent variable \(x_{ij}^k\) such
that they have a mean of zero and a standard deviation of one by cluster-mean
centering with clusters being an individual participant, as recommended by
\cite{Enders2007}. The clusters are chosen as all data from an individual
participant because we are only interested in the association between the state
of the balance assist system and the outcome of the perturbation for that
participant, not how participants perform relative to each other. We expected
there to be variation between participants in how well they are able to resist
the perturbation; however, this was not true. If a random effect is included in
the model for each participant, the intraclass correlation (ICC) is less that
3\% for both the 6~\si{\kph} and 10~\si{\kph} datasets, so we utilize a simple
single-level logistic regression model without the random effect instead of a
multilevel model. This tracks with the independent data collected in
\cite{Haitjema2023a} which also had negligible participant variation. This left
us with angular impulse, perturbation order, balance assist state, and roll \&
steer angles at the time of perturbation as independent variables. We also
include interaction effects between the balance assist state and the other four
variables. We divide the analysis into two separate model evaluations, one for
the 6~\si{\kph}, \(\kappa=5.2\) trials and one for the 10~\si{\kph},
\(\kappa=3.9\) trials and we evaluate our hypothesis for each set of data.
Statistics were computed with R version 4.1.3~\cite{RCoreTeam2021} the lme4
version 1.1~\cite{Bates2015} package.

@@

\subsection{Stability and Human Controlled Plant Dynamics}
%
The linear Carvallo-Whipple model indicates that the steer controller
stabilizes the bicycle-rider system, but this model assumes the rider's hands
are not connected to the handlebars and that they clamp their body as rigidly
as possible to the rear frame. In reality, the system's behavior is likely more
akin to a marginally stable or an easily controllable unstable system due to
the various un-modeled effects. Our system may not result in a definitely
stable system, i.e. cannot fall, but having plant eigenvalues with very small
unstable eigenvalue real parts correlates to ease of control~\cite{Hess2012}
which can be beneficial for avoiding falls.

The controller design we utilize, Equation~\ref{eq:implemented-controller},
also increases the weave mode frequency by a factor of about three in a
bandwidth of about 1~\si{\hertz}. This bandwidth is still controllable by the
human's neuromuscular system~\cite{Magdaleno1971}, but may feel unusual as it
is more akin to what the steering would feel like at in the
\SIrange{30}{40}{\kph} speed range for a normal bicycle. Hanakam, Wehner, and
Wrede~\cite{Hanakam2023} reported dissatisfaction in subjective rider feeling
on their similar bicycle to ours and this high frequency effect to the
human-controlled plant dynamics could be an explanation to their findings.
